Outline flowCore paper

### 2008-05-05 paper/manuscript.tex
small work on the text and add references
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31503 bc3139a8-67e5-0310-9ffc-ced21a209358

\section*{Introduction}
Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry high content screening (FC-HCS) have recently been
developed and are now used in both basic and clinical research. As a result, the need for flexible and well structured computational tools to
efficiently handle and analyze FC-HCS has increased
considerably. 
%Indeed, the amount of information generated by these technologies must be stored, managed, and needs to be summarized in order to be accessible to the researcher. This sentance is not grammaically corect, and really doesn't add anything anyway
However, the absence of a
shared research platform that bioinformaticians, computer scientists,
and statisticians can utilize to develop novel %or standard - not sure we want to advocate standard methods? Do we really want to say everyone should use the same method for analysis?
methods for
flow cytometric analysis has hindered innovation. <ref Lizard G. FCM analyses and bioinformatics: Interest in new softwares to
optimize novel technologies and to favor the emergence of innovative concepts in cell
research. Cytometry A 2007;71A:646–647>
We believe that the
open source statistical software R, in conjunction with the
Bioconductor Project can fill this gap.  In this paper, we propose R
data structures to handle FCM data through the main steps
of importing, storing, assessing and preprocessing. These structures,
along with a range of specialized methods for compensation,
transformation, gating and additional preprocessing steps, are
implemented in the newly developed Bioconductor package flowCore.  flowCore is a freely available, sophisticated and extensible FCM data analysis platform that enables researchers to
efficiently handle FC-HCS data and enables open development of new
methods and tools for their coherent analysis. flowCore runs on
Windows, Mac OS X and Linux/Unix operating systems.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
platform that bioinformaticians, computer scientists, and
statisticians routinely use to develop novel tools and methodology for
statistical data analysis. For biological applications, the
Bioconductor project has become the standard toolset
\citep{gentleman2006bos} to process data spanning a diversity of
research fields from genomics to proteomics to complex cell
biology. It is particularly geared towards the analysis of large
numeric datasets that typically arise from high throughput experiments
such as microarray gene expression analysis or FC-HCS. A shared
platform proved to be particularly beneficial for the development of
analysis routines for microarray gene expression data, and by today
many of these early innovations have been incorporated into commercial
software products, thus greatly improving the quality of gene
expression data analysis. Shared data structures have been shown to be
essential for such a collaborative effort. We envision a similar
paradigm for FC-HCS data.

\subsection*{Existing data standards and conventions}
Currently, data from FCM experiments are stored in single
files according to the Flow Cytometry Standard (FCS)<add ref to FCS3 Cytometry article>. 
%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup. - I believe FCS only tries to capture a subset of meta information
However, the recent
developments in high-throughput FCM are shifting the focus
of interest away from single-tube based measurements towards large and
complex experimental designs with dozens of co-variates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of %most of <can you think of an example that is not closed source?>
these products makes it impractical to
connect to and integrate the data into an analysis pipeline.

Traditionally, the majority of FCM experiments have been
analyzed manually by direct data inspection in one or two dimensions,
or by very basic comparisons of summary statistics. We believe that
these approaches do not fully address the highly complex nature of FCM
data, in particular, they discard many of the fundamental aspects of
the data, such as its underlying distribution and its high-dimensional
nature. In addition, the subjective character of manual analyses are a
major obstacle to reproducibility. For FC-HCS data, unassisted manual
inspection has become time and labor intensive, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that close collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's
high-throughput paradigm.

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
represents a proposal on how to form unambiguous XML-based gate
definitions, which can facilitate the interchange and validation of
data between different software packages with the potential of
significant increase of hardware and software interoperability. Gates
may be ordered into a hierarchical structure to describe a gating
strategy. They may be applied on parameters as in FCS files or on
transformed parameters as described by an explicit parameter
transformation. This enables the encoding of a simple analytical work
flow, in a way that allows us to reconstruct the analysis
programmatically.

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementation of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing work flows and to communicate with
 every other FCM tool that adheres to the proposed
standard.

@@

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors that occur when using more or less ad hoc spreadsheets that lack a
controlled structure. In general, there is no native ordering to a
flow set, which in a way reflects the still predominant application of
tube-based experiments rather than experiments following a
plate-design. However, the flow set structure can be readily extended
to also accommodate such spatial information \textit{(plateCore
  reference, Perry, Errol?)}.

@@

\section*{Standard flow operations}
The basic operations in all FCM analyses are typically the
same: the data need to be compensated (if not already done on the
instrument side) and transformed, and sup-populations of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for FCM analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven interface. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flow frames as well as on entire flow
sets. While transformation, and to a certain extent compensation,
are fairly routine operations with only limited potential for
improvement, being able to implement new methodologies for gating of
FCM data, and extending the capabilities of flowCore
through object oriented programming are features that clearly sets our
framework apart from other FCM analysis tools. By factoring out as
much of the bookkeeping as possible, programmers can focus on the
actual operations rather then having to deal with the tedious details
of data integration and access. Third-party methods can act on their own as first-class citizens in the analysis
framework, without breaking the workflow or the basic
infrastructure. This design allows for the straightforward extension
of flowCore's capabilities, and has already fostered the development of a
number of valuable add-ons. \citep{lo2008agf,sarkar2008ufv}
\textit{(plateCore reference, Perry, Errol?)}.

\subsection*{Transformation and compensation}
Data transformation is essential for both data
visualization and modelling \citep{lo2008agf}. The major
transformations that are routinely used in FCM analysis
have been implemented in flowCore <e.g., subset list?>. Furthermore, the design of the R
language makes it easy to define arbitrary functions to apply to the
data of individual flow frames or entire flow sets,
respectively. Compensation is available for both flow frames and flow
sets, and the software also offers functionality to compute spillover
or compensation matrices from a set of appropriate compensation
samples.

\subsection*{Gating}
In flowCore, gating operations are represented by classes that can be
extended in an object-oriented manner. Basic gate types such as
rectangular gates, ellipses and polygon gates are implemented as part
of the framework. In addition, we introduce the notion of data-driven
gates, or filters, for which the necessary parameters are computed
based on the properties of the underlying data, for instance by
modelling underlying data distribution or by density estimation. The
ability to programmatically access gates is a prerequisite for
automated or semi-automated gating. By utilizing a unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, like the proportions of
events falling in single gates or in the combinations of multiple
gates. Complex combinations and hierarchies of gates can be captured
in objects of class filterSet, allowing to apply multi-step gating
strategies. The definition of gates in flowCore follows the Gating-ML
CR, thus any flowCore gating strategy can be reproduced by any other
software that also adheres to the standard.

@@

%%not sure about this next section vs. the use of Suuplementary material. Would it be useful/informative to show the code required to generate the plot? Otherwise I'm not sold that the next paragraph adds anything.

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flow frame or a complete flow set
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarily, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

@@

Static gating for all samples in a high-throughput FCM
experiment is often not an option, since the measured variables tend
to vary between different treatments or over time. Automated or
data-driven gating tries to estimate the gating regions from the
underlying data, thus providing a fast objective solution to the
analysis of potentially very large and diverse data sets
\cite{lo2008agf}. One of the automated gating methods implemented in
flowCore is based on identifying areas of significant curvature in a
kernel density estimate of the data \citep{wand2008}. Assuming that
the regions of interest are of high density, the software is able to
reliably detect them in either the one ore the two-dimensional density
landscape. Kernel density estimation is a well-known problem in
statistical computing, and a lot of effort has been invested in the
development of good software to address it. The modular design of R,
Bioconductor and flowCore allows to easily integrate these existing
solutions into our framework. In this example, we directly use R code
from the feature package developed by Wand et al. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionality
implemented in flowCore is available to other package authors, as we
have seen with the afore-mentioned flowClust package.

Through flowCore, we have provided the FCM community a freely available, sophisticated, development and
analysis platform for high throughput data analysis. The scope and complexity of
the technology is evolving rapidly, and a collaborative effort in
devising new methodolgy has been proven beneficial for a number of
different biological and computational biology challenges. We hope
that our framework will be the foundation for fruitful shared research
by many calloborators from multiple scientific fields, further pushing
the boundaries of the exiting and still promising FCM
technology.

>>>>

\section*{Introduction}
Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry high content screening (FC-HCS)
have recently been developed and are now used in both basic and
clinical research. As a result, the need for flexible and well
structured computational tools to efficiently handle and analyze
FC-HCS has increased considerably.
%Indeed, the amount of information generated by these technologies must be stored, managed, and needs to be summarized in order to be accessible to the researcher. This sentance is not grammaically corect, and really doesn't add anything anyway
However, the absence of a shared research platform that
bioinformaticians, computer scientists, and statisticians can utilize
to develop novel \ref{lizard2007fca}.
% or standard - not sure we want to advocate standard
% methods? Do we really want to say everyone should use the same method
% for analysis?  methods for flow cytometric analysis has hindered
% innovation. <ref Lizard G. FCM analyses and bioinformatics: Interest
% in new softwares to optimize novel technologies and to favor the
% emergence of innovative concepts in cell research. Cytometry A
% 2007;71A:646–647> \ref{lizard2007fca}
We believe that the open source statistical
software R, in conjunction with the Bioconductor Project can fill this
gap.  In this paper, we propose R data structures to handle FCM data
through the main steps of importing, storing, assessing and
preprocessing. These structures, along with a range of specialized
methods for compensation, transformation, gating and additional
preprocessing steps, are implemented in the newly developed
Bioconductor package flowCore.  flowCore is a freely available,
sophisticated and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and enables open
development of new methods and tools for their coherent
analysis. flowCore runs on Windows, Mac OS X and Linux/Unix operating
systems.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
platform that bioinformaticians, computer scientists, and
statisticians routinely use to develop novel tools and methodology for
statistical data analysis. For biological applications, the
Bioconductor project has become the standard toolset
\citep{gentleman2006bos} to process data spanning a diversity of
research fields from genomics to proteomics to complex cell
biology. It is particularly geared towards the analysis of large
numeric datasets that typically arise from high throughput experiments
such as microarray gene expression analysis or FC-HCS. A shared
platform proved to be particularly beneficial for the development of
analysis routines for microarray gene expression data, and by today
many of these early innovations have been incorporated into commercial
software products, thus greatly improving the quality of gene
expression data analysis. 
%% NLM: that's probably the important point: platform for innovation rather than for standardization
Shared data structures have been shown to be
essential for such a collaborative effort. We envision a similar
paradigm for FC-HCS data.

\subsection*{Existing data standards and conventions}
Currently, data from FCM experiments are stored in single files
according to the Flow Cytometry Standard (FCS) \ref{seamer1997pnd}.
%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup. - I believe FCS only tries to capture a subset of meta information
However, the recent
developments in high-throughput FCM are shifting the focus
of interest away from single-tube based measurements towards large and
complex experimental designs with dozens of co-variates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of %most of <can you think of an example that is not closed source?>
these products makes it impractical to
connect to and integrate the data into an analysis pipeline.

Traditionally, the majority of FCM experiments have been analyzed
manually by direct data inspection in one or two dimensions, or by
very basic comparisons of summary statistics. We believe that these
approaches do not fully address the highly complex nature of FCM data,
in particular, they discard many of the fundamental aspects of the
data, such as its underlying distribution and its high-dimensional
nature. In addition, the subjective character of manual analyses are a
major obstacle to reproducibility. For FC-HCS data, unassisted manual
inspection has become time consuming and labor intensive, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that close collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's high-throughput
paradigm.

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
represents a proposal on how to form unambiguous XML-based gate
definitions, which can facilitate the interchange and validation of
data between different software packages with the potential of
significant increase of hardware and software interoperability. Gates
may be ordered into a hierarchical structure to describe a gating
strategy. They may be applied on parameters as in FCS files or on
transformed parameters as described by an explicit parameter
transformation. 
%% I find the previous sentence redundant - 
%% They may be applied on raw or transformed parameters
This enables the encoding of a simple analytical workflow, in a way
that allows us to reconstruct the analysis programmatically.

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementation of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing workflows and to communicate with
 every other FCM tool that adheres to the proposed
standard.

@@

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
that occur when using more or less ad hoc spreadsheets that
lack a controlled structure. There is no native ordering
to a flow set, which in a way reflects the still predominant
application of tube-based experiments rather than experiments
following a plate-design. However, the flow set structure can be
readily extended to also accommodate such spatial information
\textit{(plateCore reference, Perry, Errol?)}.

@@

\section*{Standard flow operations}
The basic operations in FCM analyses are typically the same: the data
need to be compensated (if not already done on the instrument side)
and transformed, and sup-populations of interest need to be selected
based on a set of (predominantly sequential) gates. All software
solutions for FCM analysis offer more or less elaborate support for
these kinds of operation, and most of them in an interactive,
GUI-driven interface. The approach we have taken in flowCore is to
abstractly describe these operations and build a set of tools to
perform them on both flow frames and flow sets. While transformation,
and to a certain extent compensation, are fairly routine operations
with only limited potential for improvement, being able to implement
new methodologies for gating of FCM data, and extending the
capabilities of flowCore through object oriented programming are
features that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the workflow or
the basic infrastructure. This design allows for the straightforward
extension of flowCore's capabilities, and has already fostered the
development of a number of valuable add-ons
\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore reference, Perry,
Errol?)}.

\subsection*{Transformation and compensation}
Data transformation is essential for both data visualization and
modelling \citep{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in flowCore
(\textit{e.g.,} quadratic, log or arcsinh). Furthermore, the design of the R language makes
it easy to define arbitrary functions to apply to the data of
individual flow frames or entire flow sets, respectively. Compensation
is available for both flow frames and flow sets, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate compensation samples.

\subsection*{Gating}
In flowCore, gating operations are represented by classes that can be
extended in an object-oriented manner. Basic gate types such as
rectangular gates, ellipses and polygon gates are implemented as part
of the framework. In addition, we introduce the notion of data-driven
gates, or filters, for which the necessary parameters are computed
based on the properties of the underlying data, for instance by
modelling underlying data distribution or by density estimation. The
ability to programmatically access gates is a prerequisite for
automated or semi-automated gating. By utilizing a unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, like the proportions of
events falling in single gates or in the combinations of
gates. Complex combinations and hierarchies of gates can be captured
in objects of class filterSet, allowing to apply multi-step gating
strategies. The definition of gates in flowCore follows the Gating-ML
CR, thus any flowCore gating strategy can be reproduced by any other
software that also adheres to the standard.

@@

%% not sure about this next section vs. the use of Suuplementary material. Would it be useful/informative to show the code required to generate the plot? Otherwise I'm not sold that the next paragraph adds anything.

%% I am not sure about the use of Supplementary Material - flowCore and all the packages mentioned here 
%% already have at least 1 vignette with examples and code

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flow frame or a complete flow set
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarly, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

@@

Static gating for all samples in a high-throughput FCM experiment is
often not an option, since the measured variables tend to vary between
different treatments or over time. Automated or data-driven gating
tries to estimate the gating regions from the underlying data, thus
providing a fast objective solution to the analysis of potentially
very large and diverse data sets \cite{lo2008agf}. One of the
automated gating methods implemented in flowCore is based on
identifying areas of significant curvature in a kernel density
estimate of the data \citep{wand2008}. Assuming that the regions of
interest are of high density, the software is able to reliably detect
them in either the one or the two-dimensional density
landscape. Kernel density estimation is a well-known problem in
statistical computing, and a lot of effort has been invested in the
development of good software to address it. The modular design of R,
Bioconductor and flowCore allows to easily integrate these existing
solutions into our framework. In this example, we directly use R code
from the feature package developed by \cite{wand2008}. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionality
implemented in flowCore is available to other package authors, as we
have seen with the afore-mentioned flowClust package.

Through flowCore, we have provided the FCM community a freely
available, sophisticated, development and analysis platform for high
throughput data analysis. The scope and complexity of the technology
is evolving rapidly, and a collaborative effort in devising new
methodology has been proven beneficial for a number of different
biological and computational biology challenges. We hope that our
framework will be the foundation for fruitful shared research by many
collaborators from multiple scientific fields, further pushing the
boundaries of the exiting and still promising FCM technology.
